---
title: "Arithmetic with Date-Time Objects"
author: "Siddharth"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE,warning=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Like number Arithmetic, date-time objects can also be added, subtracted, multiplied and divided. For e.g.

\- You might want to subtract dates from each other to find out length of interval between two dates.

\- You might want to add the dates together to generate a new date in future.

\- You might want to divide one time span by another, to work out the number of weeks in a certain interval.

## Load the library

```{r}
library(tidyverse)
```

## Simple arithmetic with Dates - Subtraction

```{r}
launch_date <- dmy_hm("14072023 09:05")

land_date <- dmy_hm("23082023 12:33")

time_travel <- land_date - launch_date

time_travel
```

## Simple arithmetic with Dates - Division

```{r}
time_travel_weeks <- time_travel/7

time_travel_weeks

```

## The Base R function - difftime()

When we subtract one date from another, at the back-end R uses the `difftime()` function to calculate the difference.

```{r}
time_travel_weeks <- difftime(land_date,launch_date,units="weeks")

time_travel_weeks

```

## Let's work on billing transactions data and use date-time arithmetic.

```{r}
transact <- read.csv('transact.csv')
head(transact)
```

## Let's answer the questions:

-   How many days are required to settle the Invoice?
-   How many invoices were settled post due date?

### How many days are required to settle the Invoice?

```{r}
class(transact$Invoice)
class(transact$Due)
class(transact$Payment)
```

```{r}
transact %>%
  mutate(days_to_pay = difftime(ymd(Payment),ymd(Invoice),units="days"))%>%
 head()
```

### How many invoices were settled post due date?

```{r}
transact %>%
  mutate(delay = difftime(ymd(Payment),ymd(Due),units="days"))%>%
  filter(delay>0)%>%
  count(delay)%>%
 head()
```

## Date-time objects can be worked with comparison operators.

### For Infosys Ltd, compute the number of shares traded and its value in Quarter1 of the financial year.

```{r}
share_data <- read.csv("share_price.csv")
glimpse(share_data)
```

```{r}
share_data %>%
  mutate(Date = dmy(Date))%>%
  filter((Date > ymd("20220331"))&(Date < ymd("20220701")))%>%
  summarise(total_volume = sum(No.of.Shares),
            total_value = sum(`Total.Turnover..Rs..`))%>%
  head()
```

## Concept of time span

In `lubridate` package, there are two separate classes of time span. A `Duration` class and `Period` Class.

All the helper functions for creating time-span are named after the units of time. Please note they are all in plural form.

The helper functions for Duration start with "d". For period there is no change.

| Time span | Duration   | Period    |
|-----------|------------|-----------|
| years     | `dyears()` | `years()` |
| days      | `ddays()`  | `days()`  |
| weeks     | `dweeks()` | `weeks()` |
| hours     | `dhours()` | `hours()` |

## Duration

`Duration` is a fixed time period. Durations are not influenced by leap years or other time changes. It adds/subtracts fixed number of seconds to the current date to calculate the future or past date.

```{r}
date1 <- dmy("01012024")
date1 + dyears(1)
```

```{r}
leap_year(2024)
```

## Period

`Period` uses human concept of time span. Periods will take into account the leap year, day light saving time and other time related subtleties and produce the result.

```{r}
date1 <- dmy("01012024")
date1 + years(1)
```

## Simple Arithmetic with `Period` and `Duration`

```{r}
days()
```

```{r}
2*days()
days(2)
```

```{r}
ddays(2)
```

```{r}
ymd("2011-01-01") + days()
```

## Summary:

-   We can perform arithmetic with date-times. We can add and subtract dates, perform division and multiplication. We can also use comparison operators on date-times.

-   We carried out some arithmetic operations on `transactions` data set. We also used comparison operators to filter out the Quarter1 time range from `share_price` dataset.

-   `lubridate` has two time-span classes - `Duration` class and `Period` class

-   `Duration` class follows stop-watch concept of time-span. A duration year will always equal 365 days. Thus, duration class won't take into account leap years or any other time changes.

-   `Period` class follows human concept of time span. Periods will take care of leap year, day-light saving calculations and generate the result.

-   Period and Duration have the same helper functions. All these helper functions are named after period of time in plural form. However, for duration, these functions start with prefix "d".

-   All arithmetic and comparison operations can be performed on Duration and Period classes.
