---
title: "Extract information from date-time objects"
author: "Siddharth"
date: "`r Sys.Date()`"
output: html_document
---

In this video we will have a look at various functions from `lubridate` package, that are used to extract the information from date-time objects. For e.g. we can use `year()` function to extract year value from date-time object.

## Load the library

```{r}
library(tidyverse)
```

## Generate simulated time series data

```{r}
set.seed(123)

data_time <- data.frame(date = sample(seq(as.POSIXct("2020/01/01 03:12:34",tz=" "),
                                     by = "day",
                                     length.out = 100),
                                     replace = TRUE),
                   value = round(rnorm(100, 5, 2), 2))

head(data_time)
```

## Extract year from date-time object

```{r}
data_time %>%
  mutate(year_col=year(date))%>%
  head
```

## Extract month from date-time object

```{r}
data_time %>%
  mutate(year_col=year(date),month_col=month(date))%>%
  head

```

## Extract month from date-time object and assign the `label` to it.

```{r}
data_time %>%
  mutate(year_col=year(date),month_col=month(date,label=TRUE))%>%
  head
```

## Extract hour from date-time object

```{r}
data_time %>%
  mutate(year_col=year(date),month_col=month(date),hour_col=hour(date))%>%
  head
```

Let's assume, we are analyzing Delta Airlines' performance in terms of arrival delay. So here are the questions we are interested in looking at.

-   On average, Which month has the largest arrival delay?

-   On average, Which week has the largest arrival delay?

-   On average, Which day of a week has the largest arrival delay?

## Load the library

```{r}
library(nycflights13)
glimpse(flights)
```

## On average, Which month has the largest arrival delay?

### Extract month from a date

```{r}
data_flights <- flights %>%
  select(time_hour,arr_delay,carrier)%>%
  filter(carrier=="DL")%>%
  mutate(data_month = month(time_hour,label=TRUE))
head(data_flights)
```

### Aggregate `arrival delay` by month

```{r}
data_flights_aggr <- data_flights %>%                         
  group_by(data_month) %>% 
  dplyr::summarize(arr_delay = mean(arr_delay,na.rm = TRUE)) %>% 
  as.data.frame()
```

### Visualise `arrival delay` by month

```{r}
data_flights_aggr %>%
  ggplot(aes(x=data_month,y=arr_delay))+
  geom_col(fill="khaki",color="blue")+
   geom_text(aes(label=round(arr_delay,0)),nudge_y = 1)+
  xlab("Month")+
  ylab("Arrival Delay (Minutes)")+
  theme_classic()
```

What observations can we make on arrival delays of Delta Airlines?

-   When compared to other months, June and July have the largest arrival delays.

-   The average delay in June and July is 13 minutes and 15 minutes, respectively.

-   What do these negative values tell us? They are negative because flights arrived before their scheduled arrival times.

## On average, Which day of a week has the largest arrival delay?

### Extract day from a date

```{r}
data_flights_week <- data_flights %>%
    mutate(data_day = wday(time_hour,label=TRUE)) 
head(data_flights_week)
```

### Aggregate `arrival delay` by days

```{r}
data_flights_aggr_week <- data_flights_week %>%                         
  group_by(data_day) %>% 
  dplyr::summarize(arr_delay = mean(arr_delay,na.rm = TRUE)) %>% 
  as.data.frame()
```

### Visualise `arrival delay` by days

```{r}
data_flights_aggr_week %>%
  ggplot(aes(x=data_day,y=arr_delay))+
  geom_col(fill="khaki",color="blue")+
  geom_text(aes(label=round(arr_delay,0)),nudge_y = 1)+
  xlab("Day")+
  ylab("Arrival Delay (Minutes)")+
  theme_classic()
```

What observations can we make on arrival delays of Delta Airlines?

-   Mondays and Thursdays have the largest arrival delays compared to other days.

-   the average delay on Monday and Thursday is 5 minutes and 6 minutes respectively.

-   Saturday is an ideal day to travel by Delta Airlines, because you can arrive 5 minutes earlier than planned arrival time.

## On average, Which hour of a day has the largest arrival delay?

### Extract hour from a date

```{r}
data_flights_hour <- data_flights %>%
   mutate(data_hour = hour(time_hour))
head(data_flights_hour)
```

### Aggregate `arrival delay` by hours

```{r}
data_flights_hour_aggr <- data_flights_hour %>% 
  group_by(data_hour) %>% 
  dplyr::summarize(arr_delay = mean(arr_delay,na.rm = TRUE)) %>% 
  as.data.frame()
```

### Visualise `arrival delay` by hour

```{r}
data_flights_hour_aggr %>%
  ggplot(aes(x=data_hour,y=arr_delay))+
  geom_col(fill="khaki",color="blue")+
  geom_text(aes(label=round(arr_delay,0)),nudge_y = 1)+
  xlab("Hour")+
  ylab("Arrival Delay (Minutes)")+
  scale_x_continuous(breaks=c(1:24))+
  theme_classic()
```

What observations can we make on arrival delays of Delta Airlines?

-   From 6AM until 10AM there are no delays and flights have arrived ahead of schedule.

-   The arrival delays starts to go up from 11AM and they are highest at 9PM.

-   On average you can save 11 minutes on your commute, if you travel at around 11PM.

-   There are no Delta Airlines flights scheduled from 12AM to 5AM.

## Summary:

-   `lubridate` package has various functions to extract the information from date-time objects.

| function                      | output |
|-------------------------------|--------|
| year('2020-01-31 03:12:34' )  | 2020   |
| month('2020-01-31 03:12:34' ) | 01     |
| day('2020-01-31 03:12:34')    | 31     |
| hour('2020-01-31 03:12:34')   | 03     |
