---
title: "Time intervals"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning = FALSE)
```

Interval is a duration with specific start and end date. `lubridate` package comes with many helper functions that work efficiently with interval.

## Load the library

```{r}
library(tidyverse)
```

## Load the data

```{r}
india_PM <- read.csv("India_prime-ministers.csv")
    head(india_PM)
```

## Create the interval

```{r}
class(india_PM$From)
class(india_PM$To)

india_PM <- india_PM %>% 
      mutate(office_interval = interval(dmy(From),dmy(To)))%>%
  view

#india_PM <- india_PM %>% 
#  mutate(office_interval = dmy(From) %--% dmy(To))%>%
#  view
```

## Extract start date and end date of the interval

```{r}
 india_PM %>% 
        mutate(office_start = int_start(office_interval),
               office_end = int_end(office_interval))%>%
      view
```

## Calculate interval length

```{r}
    india_PM %>%
      mutate(office_period = int_length(office_interval))%>%
      view
```

## Calculate interval length in `Period`

```{r}
 india_PM %>%
      mutate(office_period = as.period(office_interval))%>%
      view
```

## Calculate interval length in `Duration`

```{r}
  india_PM %>%
      mutate(int_duration = as.duration(office_interval))%>%
      view
```

## Comparing intervals

Check who was the incumbent prime minister when these events took place.

```{r}
    india_world_cup_cricket <-  dmy("25 June 1983")
    india_economic_reforms <- dmy("24 July 1991")
    india_kargil_war <- interval(dmy("3 May 1996"),dmy("26 July 1996"))
```

### Who was the incumbent PM when India won the world Cup?

```{r}
india_PM[india_world_cup_cricket %within% india_PM$office_interval,1:3]
```

### Who was the incumbent PM when India ushered into the era of economic reforms?

```{r}
india_PM[india_economic_reforms %within% india_PM$office_interval,1:3]
```

### Who was the incumbent PM when Kargil war was fought between Indian and Pakistan?

```{r}
india_PM[int_overlaps(india_PM$office_interval,india_kargil_war),1:3]
```

## Summary

-   Interval is a duration with specific start and end date. `lubridate` package comes with many helper functions to work with intervals.

-   To create the interval, use `interval()` function and pass on start date and end date of the interval. The interval can also be created using `%--%`operator.

-   `int_start()` and `int_end()` functions are used to extract start date and end date of the interval.

-   `int_period()` will return interval length in periods and `int_duration()` will return the interval length in duration (in seconds)

-   To compare a datetime to an interval we used `%within%` operator. It will return `TRUE` if date-time is inside the interval and `FALSE` otherwise.

-   To compare two intervals, We have used `int_interval()` function which will return `TRUE` if there are any times that are inside both intervals.
