---
title: "Time series introduction"
author: "Siddharth"
date: "`r Sys.Date()`"
output: html_document
---

## What is Time Series?

A time series is a set of observed values ordered in time, or we can say, repeated measurement of something usually with the same fixed interval of time (hourly, weekly, monthly).

## Plot time series - Closing prices of European Stock indices

```{r}

library(MASS)
plot(EuStockMarkets)

```

## Plot time series - Monthly Airline passenger number

```{r}

plot(AirPassengers)

```

In R, there are special time series class - called as `Date` class and `POSIXct` class which are designed to work best with many time series statistical models and visualization libraries such as ggplot2.

Widely used packages for time series data manipulation:

-   `lubridate`

-   `zoo`

## Load the package

```{r}

library(tidyverse)

#Separately load the package if you have older version of tidyverse package.  

# library(lubricate) 

```

## Import the share price data

```{r}
share_infy <- read.csv("share_price.csv")
head(share_infy)
```

## check the class of time variable in data set

```{r}
class(share_infy$Date)
```

## Parsing Date

+--------------------------------------+-----------------------------------+
| order of year,month, day in data-set | parsing function in `lubridate()` |
+======================================+===================================+
| 16-February-2024                     | `dmy()`                           |
|                                      |                                   |
| day-month-year                       |                                   |
+--------------------------------------+-----------------------------------+

```{r}
share_infy_time <- share_infy %>%
  mutate(Date = dmy(Date))
```

## check the class of time variable

```{r}
class(share_infy_time$Date)
head(share_infy_time$Date)
```

## Import IoT sensor data

```{r}
room_temp <- read.csv("temp_data.csv")
head(room_temp)
```

## Check the class of time variable in data set

```{r}

class(room_temp$Datetime)

```

## Parsing Date

+--------------------------------------+-----------------------------------+
| order of year,month, day in data-set | parsing function in `lubridate()` |
+======================================+===================================+
| 2022-01-04 00:00:00                  | `ymd_hms()`                       |
|                                      |                                   |
| year-month-day-hour-minute-second    |                                   |
+--------------------------------------+-----------------------------------+

```{r}
room_temp_time <- room_temp %>%
  mutate(Datetime = ymd_hms(Datetime))
```

## check the class of time variable

```{r}
class(room_temp_time$Datetime)
head(room_temp_time$Datetime)
```

## Date format without separator

```{r}
date_1 <- ("20240322")

date_1_time <- ymd(date_1)
class(date_1_time)
date_1_time
```

## forward slash separator

```{r}
date_2 <- ("23/03/2023")

date_2_time <- dmy(date_2)
class(date_2_time)
date_2_time
```

## Another separator

```{r}
date_3 <- "2023 March 23 03:45"

date_3_time <- ymd_hm(date_3)
class(date_3_time)
date_3_time
```

## Summary:

-   A time series is a set of observed values ordered in time
-   In R, there are special time series objects - called as Date-time objects which are designed to work best with many time series statistical models and visualization libraries such as ggplot2.
-   Before manipulating time series data, we must ensure the time component of the data set is converted into date-time object that R understands.
-   there are two classes in R `Date` and `POSIXct` which are exclusively designed to handle date-time objects.
-   We used various combinations of `dmy()` function from `lubridate` package to parse the dates and times.
